---
title: "DNA"
author: "Freddy"
date: "2025-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages
```{r}

pacman::p_load(tidyverse, dplyr, purrr, tidyr, rstan, cmdstanr, tidybayes, posterior, patchwork, igraph, ggraph, dplyr, pheatmap)
```




### Load Data
```{r}

trial_df <- read_csv("trials.csv")
#distance_mat <- dist_matrix

trial_df <- trial_df %>% 
  drop_na(gadm.adm1)

trial_activity <- trial_df %>%
  group_by(region = gadm.adm0, decade) %>%
  summarise(trials = sum(tried, na.rm = TRUE)) %>%
  mutate(trial_occurred = as.integer(trials > 0)) %>%
  ungroup()

```

### Read in distance matrix
```{r}
distance_matrix <- read_csv("dist_country.csv")
# distance_matrix: matrix where 1 = neighboring
get_neighbors <- function(region_name) {
  if (!region_name %in% rownames(distance_matrix)) return(character(0))
  neighbors <- distance_matrix[region_name, ]
  names(neighbors[neighbors == 1])
}

```





```{r}
trial_activity <- trial_df %>%
  group_by(region = gadm.adm0, decade) %>%
  summarise(trials = sum(tried, na.rm = TRUE)) %>%
  ungroup()

trial_activity_country <- trial_df %>% 
  group_by(country = gadm.adm0, decade) %>% 
  summarise(trials = sum(tried, na.rm = TRUE)) %>% 
  ungroup()


# Number of regions (R) and time periods (T)
R <- length(unique(trial_activity$region))
T <- length(unique(trial_activity$decade))

# Prepare trial_occurred matrix (R x T)
trial_matrix <- trial_activity %>%
  spread(key = decade, value = trials, fill = 0) %>%
  select(-region) %>%
  as.matrix()

# List of neighbors based on  distance matrix
neighbor_ids <- list()
N_neighbors <- rep(0, R)


for (r in 1:R) {
  neighbors <- which(distance_matrix[r, ] == 1)
  neighbor_ids[[r]] <- neighbors
  N_neighbors[r] <- length(neighbors)
}

```




```{r}
# Get the number of regions 
n_regions <- length(unique(trial_df$gadm.adm1))
# Get the number of decades
n_decades <- length(unique(trial_df$decade))


```



```{r}
# Create a matrix of trial counts
trial_count_matrix <- trial_df %>%
  group_by(gadm.adm0, decade) %>%
  summarise(tried = sum(tried, na.rm = TRUE)) %>%
  spread(key = decade, value = tried, fill = 0) %>%
  as.matrix()

```


```{r}
# Number of neighbors for each region
n_neighbors <- rowSums(distance_matrix == 1)  # Sum of 1's in each row to count neighbors

# Neighboring region IDs 
neighbor_ids_matrix <- apply(distance_matrix, 1, function(x) which(x == 1))  #  neighboring regions

```



```{r}
# Remove any region from the trial count matrix that has only NA values
valid_regions <- rowSums(is.na(trial_count_matrix)) != ncol(trial_count_matrix)

trial_count_matrix <- trial_count_matrix[valid_regions, ]
neighbor_ids_matrix <- neighbor_ids_matrix[valid_regions, , drop = FALSE]
n_neighbors <- n_neighbors[valid_regions]

# Recalculate R just to check
n_regions <- nrow(trial_count_matrix)

# Check dimensions again to be sure, because of a dimension mismatch earlier
stopifnot(
  n_regions == length(n_neighbors),
  n_regions == nrow(neighbor_ids_matrix)
)

```



```{r}
# Make sure it's a numeric matrix before anything else
trial_count_matrix <- as.matrix(trial_count_matrix)
storage.mode(trial_count_matrix) <- "numeric"

# Replace any NAs with 0
trial_count_matrix[is.na(trial_count_matrix)] <- 0

trial_count_matrix <- matrix(as.integer(trial_count_matrix),
                             nrow = nrow(trial_count_matrix),
                             ncol = ncol(trial_count_matrix))



stan_data$neighbor_ids <- neighbor_ids_matrix
storage.mode(stan_data$neighbor_ids) <- "integer"
```


```{r}
# Convert tibble to data frame
distance_df <- as.data.frame(distance_matrix)

# Extract row names from the first column, then remove it
rownames(distance_df) <- distance_df[[1]]
distance_df[[1]] <- NULL


```


```{r}
# convert to a numeric matrix
distance_matrix <- as.matrix(distance_df)
storage.mode(distance_matrix) <- "numeric"


```


```{r}
# Step 6: Count neighbors for each region
n_neighbors <- rowSums(distance_matrix == 1)

# Step 7: Pad the neighbor indices
max_neighbors <- max(n_neighbors)

neighbor_ids_matrix <- t(apply(distance_matrix, 1, function(x) {
  ids <- which(x == 1)
  length(ids) <- max_neighbors  
  replace(ids, is.na(ids), 999) #for NAs
}))
storage.mode(neighbor_ids_matrix) <- "integer"
```


### Prepare stan daata
```{r}
log_trial_matrix <- log(trial_count_matrix + 1)

stan_data <- list(
  R = nrow(log_trial_matrix), 
  T = ncol(log_trial_matrix),      
  log_trials = log_trial_matrix,    
  N_neighbors = n_neighbors,          
  neighbor_ids = neighbor_ids_matrix,
  max_neighbors = max_neighbors,
  trial_count = trial_count_matrix
)

storage.mode(stan_data$neighbor_ids) <- "integer"

```



#  Run the stan model

```{r}
# Compile the Stan model
mod_log <- cmdstanr::cmdstan_model("beta_ind.stan", cpp_options = list(stan_threads = TRUE))

# Fit the model
fit_log <- mod_log$sample(
  data = stan_data,         
  seed = 1212,               
  chains = 4,               
  parallel_chains = 2,     
  threads_per_chain = 1,   
  iter_warmup = 5000,       
  iter_sampling = 5000,   
  refresh = 2000
)


print(fit_log)
```


```{r}
fit_log$summary(variables = c( "sigma_gr"))
```







### Prior-posterior update checks
```{r}

beta_post <- draws_df %>% select(matches("^beta\\["))

n_draws <- nrow(beta_post)
n_betas <- ncol(beta_post)

# Simulate prior samples 
beta_prior <- matrix(rnorm(n_draws * n_betas, 0, 1), ncol = n_betas)

# Give them the same column names as beta_post
colnames(beta_prior) <- colnames(beta_post)


```


```{r}

# Add 'source' labels
beta_post_long <- beta_post %>%
  pivot_longer(everything(), names_to = "param", values_to = "value") %>%
  mutate(source = "Posterior")

beta_prior_long <- as.data.frame(beta_prior) %>%
  pivot_longer(everything(), names_to = "param", values_to = "value") %>%
  mutate(source = "Prior")

beta_combined <- bind_rows(beta_post_long, beta_prior_long)


```



#### Prior posterior update checks
```{r}

draws_df <- as_draws_df(fit_log$draws())


set.seed(1012)  # for reproducibility
draws_df <- draws_df %>%
  mutate(
    beta_prior = rnorm(n(), mean = 0, sd = 1),
    sigma_prior = rexp(n(), rate = 1),
    sigma_gr_prior = rexp(n(), rate = 1)
  )

plot_prior_vs_posterior <- function(df, param, prior_name, label) {
  ggplot(df) +
    geom_density(aes_string(x = param), fill = "#4e825b", alpha = 0.3) +
    geom_density(aes_string(x = prior_name), fill = "#c41a80", alpha = 0.3) +
    labs(
      x = label,
      y = "Density",
      title = paste("Prior vs Posterior:", label)
    ) +
    xlim(0,4) +
    theme_bw()
}

```

```{r}
plot_prior_vs_posterior(draws_df, "beta", "beta_prior", "β (neighbor effect)")
plot_prior_vs_posterior(draws_df, "sigma", "sigma_prior", "σ (observation noise)")
plot_prior_vs_posterior(draws_df, "sigma_gr", "sigma_gr_prior", "σ_gr (regional variability)")

```








##########################################################################################

                                      Dynamic Network Analysis

#########################################################################################

```{r}
posterior_df <- as_draws_df(fit_log$draws())

# Use posterior means as point estimates
beta_hat <- mean(posterior_df$beta)
gr_hat <- posterior_df %>%
  select(starts_with("gr[")) %>%
  summarise_all(mean) %>%
  unlist()
```



```{r}

# Ensure region order matches
region_names <- trial_wide$region
T <- ncol(trial_count_matrix)
R <- nrow(trial_count_matrix)

# Initialize matrix for modeled influence
influence_matrix_list <- list()

for (t in 2:T) {
  infl_matrix <- matrix(0, nrow = R, ncol = R)
  
  for (r in 1:R) {
    neighbors <- neighbor_ids_matrix[r, 1:n_neighbors[r]]
    neighbors <- neighbors[neighbors <= R]  # Safety check
    
    for (n in neighbors) {
      # Influence of neighbor n on region r
      influence <- beta_hat * log1p(trial_count_matrix[n, t - 1])
      infl_matrix[n, r] <- influence
    }

  }
  
  influence_matrix_list[[t - 1]] <- infl_matrix
}

```



```{r}
write.csv(posterior_df, "posterior_df.csv")
```

#### Plotting the networks with igraph
```{r}
t_index <- 6

# Create graph from adjacency matrix
influence_matrix <- influence_matrix_list[[t_index]]
colnames(influence_matrix) <- rownames(influence_matrix) <- region_names
g <- graph_from_adjacency_matrix(influence_matrix, mode = "directed", weighted = TRUE, diag = FALSE)

# ---- Node attributes ----
trial_activity <- trial_count_matrix[, t_index]
V(g)$size <- rescale(trial_activity, to = c(6, 20))
V(g)$color <- "#906768"

# ---- Edge attributes ----
min_weight <- min(E(g)$weight)
E(g)$layout_weight <- E(g)$weight
if (min_weight <= 0) {
  E(g)$layout_weight <- E(g)$weight + abs(min_weight) + 1e-3
}

E(g)$color <- ifelse(E(g)$weight >= 0, "#5bb450", "#D663A9")
E(g)$alpha <- rescale(abs(E(g)$weight), to = c(0.4, 1))
E(g)$color <- alpha(E(g)$color, E(g)$alpha)

# ---- Plot ----
png("network_1360_m.png", width = 1200, height = 800) #saaavvveee
plot(g, 
     edge.width = abs(E(g)$weight) / max(abs(E(g)$weight)) * 5,
     vertex.label.cex = 2,
     vertex.label.color = "black",
     vertex.label.dist = 2,
     vertex.label.degree = pi / -1.5,
     layout = layout_with_fr(g, weights = E(g)$layout_weight))



#remember to put decade AFTER t_index, so t_index + 1
title(main = paste("Decade 1360"), cex.main = 2.5)
dev.off()
```








### Results reporting - calculating means for beta


#### beta matrix

```{r}
# Convert draws object and compute mean beta_mat
posterior_df <- as_draws_df(fit_log$draws())
R <- nrow(trial_count_matrix)

# Get posterior mean of each beta engtry in marix
beta_means <- posterior_df %>%
  select(starts_with("beta_mat[")) %>%
  summarise_all(mean) %>%
  unlist()

beta_sd <- posterior_df %>% 
  select(starts_with("beta_mat[")) %>% 
  summarise_all(sd) %>% 
  unlist

beta_summary <- posterior_df %>%
  select(starts_with("beta_mat[")) %>%
  pivot_longer(everything(), names_to = "param", values_to = "value") %>%
  group_by(param) %>%
  summarise(
    ci_lower = quantile(value, 0.025),
    ci_upper = quantile(value, 0.975),
    .groups = "drop"
  )

mean(beta_summary[["ci_lower"]])
mean(beta_summary[["ci_upper"]])
mean(beta_summary$ci_lower)


beta_mat_mean <- matrix(beta_means, nrow = R, ncol = R, byrow = TRUE)

```
```{r}
beta_offdiag <- beta_mat_mean[lower.tri(beta_mat_mean, diag = FALSE) | upper.tri(beta_mat_mean, diag = FALSE)]
mean_beta <- mean(beta_offdiag)
prop_positive <- mean(beta_offdiag > 0)


beta_offdiag
mean_beta
prop_positive
mean(beta_sd)
```


```{r}
# Get posterior mean of each growth rate
gr_means <- posterior_df %>%
  select(starts_with("gr[")) %>%
  summarise_all(mean) %>%
  unlist()

gr_sd <- posterior_df %>% 
  select(starts_with("gr[")) %>% 
  summarise_all(sd) %>% 
  unlist

gr_summary <- posterior_df %>%
  select(starts_with("gr[")) %>%
  pivot_longer(everything(), names_to = "param", values_to = "value") %>%
  group_by(param) %>%
  summarise(
    ci_lower = quantile(value, 0.025),
    ci_upper = quantile(value, 0.975),
    .groups = "drop"
  )

mean(gr_means)
mean(gr_sd)
mean(gr_summary$ci_lower)
mean(gr_summary$ci_upper)

```






### Plot for the results section - beta coefficients for countries
```{r}

range_limit <- max(abs(beta_mat_mean))  # make symmetric scale
breaks <- seq(-range_limit, range_limit, length.out = 101)

# color length matches breaks - 1, otherwise white not at neutral
colors <- colorRampPalette(c("#c41a80", "white", "#5bb450"))(100)


pheatmap(beta_mat_mean,
         cluster_rows = FALSE, cluster_cols = FALSE,
         labels_row = region_names, labels_col = region_names,
         color = colors,
         breaks = breaks)

```





### Empirical Data visualisations/Methods section


#### Look at empirical data
```{r}
#Heatmap plot

ggplot(trial_activity_country, aes(x = decade, y = reorder(country, -trials), fill = trials)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option ="plasma", direction = -1) +
  labs(title = "Heatmap of Witch Trials by Country and Decade",
       x = "Decade", y = "Modern Countries", fill = "Trials") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 7))

```





### Code for plotting the map with the countries coloured in
```{r}


# Load map data
europe_map <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf")

included_countries <- c("Germany", "France", "Poland", "Italy", "Switzerland", "Spain", "Austria", 
                        "Denmark", "Sweden", "United Kingdom", "Ireland", "Luxembourg", 
                        "Netherlands", "Belgium", "Czechia")

# Choose countries to be included
europe_map <- europe_map %>%
  mutate(included = ifelse(admin %in% included_countries, "Included", "Not Included"))


png("map.png", width = 1200, height = 800) #saaavvveee


ggplot(data = europe_map) +
  geom_sf(aes(fill = included), color = "black", size = 0.3) +
  scale_fill_manual(values = c("Included" = "#5bb450", "Not Included" = "white")) +
  coord_sf(xlim = c(-12, 35), ylim = c(35, 72), expand = FALSE) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold", margin = margin(b = 10)),
    legend.position = "none",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()  # removes weird rectangle frame
  ) +
  labs(title = "Countries Included in the Witch Trial Network Analysis")
dev.off()
```








